A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-24 text/discussion.Rmd
edit discussion of metastasis

One of the biggest hurdles in cancer research is that the search space
comprising ~20,000 protein-coding genes is comparable with the number of tumor
samples, even with multiple mutations per sample. And that is not accounting for
the fact that ~30k non-coding genes and hundreds of thousands of non-coding
regulatory elements exist with oncogenic or tumor suppressor potential
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
investigate cancer as a disease of basic cellular and biochemical pathways,
which we accomplished by translating gene-level mutations into pathway level
disruptions. Our approach differs from previously proposed methods [@PATHNET] in
that we chose to focus on pathways as the unit of disruption instead of the
gene, using the conceit that individual mutations may be sufficient to disrupt
entire pathways (as is often the case with drivers). Rather than evaluating the
burden of mutations on each pathway within or across tumors, which can be biased
by the number of genes or the coverage of pathway-linked genes in the genome,
and which we find to be relatively uniformly distributed, we scored each pathway
in binary fashion, allowing one mutation or multiple mutations to count equally
in a sample. To our knowledge, this approach has not previously been attempted
in the literature in spite of its relative simplicity.

We limited our analysis to mutations with likely deleterious effects, including
missense, nonsense, frame-shift stop-gain and splice site mutations in genes
that are actively expressed in each cancer type, thus avoiding bias from
transcription coupled repair. Our method of filtration differs from the
"rank-and-cut" method of [@PCAWG-MAIN] but represents a reasonable attempt to
account for the same biases. We also filtered genes on the basis of pathway
membership in biochemical pathways, excluding curated genesets related to
diseases, syndromes, or classes of proteins ( _e.g._ transcription factors). We
chose this approach to limit redundancy, which is already high in the remaining
pathways for some driver genes, and to exclude biologically unrelated
collections of genes. We do not account for non-coding mutations, amplification
events or structural variants, though we hope to include them in subsequent
studies (see below). Our pathway-centric analysis also allows us to circumvent
the problem that some cancer-promoting mutations likely occur at low enough
frequencies as to be indistinguishable from background. Attempts to capture
these associations better using Bayesian networks have yielded important new
insights into this problem and new genes that would not have been possible with
traditional analysis [@hornNetSigNetworkbasedDiscovery2018], supporting the case
that pathway-centric approaches like ours have the potential to increase
statistical power.

@@

We find that most tumors segregate into one of twelve biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
arm clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify cancer type
that was exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
primarily reflect differences in cell-of-origin, rather than fundamental
differences in proximal cause. It has been hypothesized that differences in the
molecular regulators of development in the array of cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes].
Consistent with this view we found that breast tumor samples of the Luminal A
subtype were heavily biased toward membership in clusters 1 and 7, and basal
tumor samples were biased toward cluster 6 (**figure 1J**). However, both
subtypes contained samples in every other class without exception, and Luminal B
and Her2 positive samples are distributed across clusters. We interpret these
data to mean that cell-of-origin molecular machinery can predispose certain cell
types within the breast epithelium to forming tumors of one class or another but
is not determinative; this view is compatible with the previously stated
hypothesis but opens the way for a more granular view of individual tumors.
Indeed, it would be surprising if we did not observe bias for some cancers and
subtypes amongst our classes, since some treatment regimens have greater
efficacy for a significant fraction of patients of a given cancer/histological
subtype. Nonetheless, the basis for some tumors being treatment-refractory in
spite of receiving the standard of clinical care for diagnostic markers remains
elusive. Doubtless some of this is due to bad luck, as tumors can metastasis and
remain dormant years before they are detected at distal sites [@hallmarks_II], or
resistant clones may have already arisen at undetectable levels, but our
analysis suggests the possibility of identifying more informative molecular,
histological or cellular subtypes that could form a basis for future separation
of patients into different precision treatment regimens.

@@

We report that pathway enrichment in metastatic tumors across all classes
(pooled) yielded generally lower effect sizes and larger p-values than the
cluster specific analysis, suggesting that signal is diluted when the samples
are analyzed this way, and supporting the view that metastasis has
class-specific requirements. Since the number of metastatic tumors is relatively
low this part of our analysis is likely underpowered and subject to expanded
analysis with larger cohorts. Nonetheless, the class-specific metastatic pathway
enrichment we observe (in **table 2**) could be a result of the pathways
directly promoting metastatic phenotypes in these classes, or because they
acquire a stronger or more complete oncogenic phenotype, since we don’t know the
order in which they were acquired in individual tumor samples. The fact that
most metastatic enrichment is cluster-specific and has a tendency to overlap
with cluster-specific pathways from non-metastatic tumors of neighboring
clusters favors the latter interpretation, and suggests a mechanism of
complementarity, in which newly acquired mutations result in similar classes
converging on a single deadly phenotype that captures most or all of the
critical features of end-stage cancer. Some clusters were characterized by
cluster-specific pathways that seem superficially unrelated; for example cluster
3 has mutations impacting protein translation and turnover, whereas cluster 5
has defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
traditional extra-cellular signaling pathways (as exemplified in **table 1**).
It may be that the primary effects of mutations in polymer metabolism in cluster
3 and 5 point to currently unappreciated mechanisms impacting hallmark pathways.
Clusters 3 and 5 are also marked by a decidedly different set of pathways in
metastasis. Clusters 3 and 5 notably have (in addition to EGFR signaling)
mutations in glucagon and glycogen metabolism, respectively, in their metastases
(**table 2**), consistent with energy metabolism as one of the new hallmarks
[@hallmarks_II], but also distinguishing these classes from the others in the
order or nature of hallmark acquisition. We submit that when we discover the
requirements of each cluster with respect to metastasis we may be able to target
them therapeutically and prevent further adaptation.

>>>>

One of the biggest hurdles in cancer research is the size of the search space;
~20,000 protein-coding genes is comparable with the number of tumor samples,
even with multiple mutations per sample. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
simplify the problem by investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously proposed
methods [@PATHNET] in that we chose to focus on the pathway as the unit of
disruption instead of the gene, employing the conceit that individual mutations
may be sufficient to disrupt entire pathways (as is often the case with
drivers). Rather than evaluating the burden of mutations on each pathway within
or across tumors, which can be biased by the number of genes or the coverage of
pathway-linked genes in the genome, and which we find to be relatively uniformly
distributed, we scored each pathway in binary fashion, allowing one mutation or
multiple mutations to count equally for each sample. To our knowledge, this
approach has not previously been attempted in the literature in spite of its
relative simplicity. We limited our analysis to mutations with likely
deleterious effects, including missense, nonsense, frame-shift, stop-gain and
splice site mutations in genes that are actively expressed in each cancer type,
thus avoiding bias from transcription coupled repair. Our method of filtration
differs from the "rank-and-cut" method of [@PCAWG-MAIN] but represents a
reasonable attempt to account for the same biases. We also filtered genes on the
basis of pathway membership in biochemical pathways, excluding curated genesets
related to diseases, syndromes, or classes of proteins ( _e.g._ transcription
factors). We chose this approach to limit redundancy, which is already high in
the remaining pathways for some driver genes, and to exclude biologically
unrelated collections of genes. We do not account for non-coding mutations,
amplification events or structural variants, though we hope to include them in
subsequent studies (see below). Our pathway-centric analysis also allows us to
circumvent the problem that some cancer-promoting mutations likely occur at low
enough frequencies as to be indistinguishable from background. Attempts to
capture these associations better using Bayesian networks have yielded important
new insights into this problem and new genes that would not have been possible
with traditional analysis [@hornNetSigNetworkbasedDiscovery2018], supporting the
case that pathway-centric approaches like ours have the potential to increase
statistical power.

@@

We find that most tumors segregate into one of twelve biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
arm clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify cancer type
that was exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
primarily reflect differences in cell-of-origin, rather than fundamental
differences in proximal cause. It has been hypothesized that differences in the
molecular regulators of development in the array of cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes].
Consistent with this view we found that breast tumor samples of the Luminal A
subtype were heavily biased toward membership in clusters 1 and 7, and basal
tumor samples were biased toward cluster 6 (**figure 1J**). However, both
subtypes contained samples in every other class without exception, and Luminal B
and Her2 positive samples are distributed across clusters. We interpret these
data to mean that cell-of-origin molecular machinery can predispose certain cell
types within the breast epithelium to forming tumors of one class or another but
is not determinative; this view is compatible with the previously stated
hypothesis but opens the way for a more granular view of individual tumors.
Indeed, it would be surprising if we did not observe bias for some cancers and
subtypes amongst our classes, since some treatment regimens have greater
efficacy for a significant fraction of patients of a given cancer/histological
subtype. Nonetheless, the basis for some tumors being treatment-refractory in
spite of receiving the standard of clinical care for diagnostic markers remains
elusive. Doubtless some of this is due to bad luck, as tumors can metastasize
and remain dormant years before they are detected at distal sites, or resistant
clones may have already arisen at undetectable levels [@hallmarks_II], but our
analysis suggests the possibility of identifying more informative molecular,
histological or cellular subtypes that could form a basis for future separation
of patients into different precision treatment regimens.

@@

We report that pathway enrichment in metastatic tumors
across all classes (pooled) yielded generally lower effect sizes and larger
p-values than the cluster specific analysis, suggesting that signal is diluted
when the samples are analyzed this way, and supporting the view that metastasis
has class-specific requirements. Since the number of metastatic tumors is
relatively low this part of our analysis is likely underpowered and subject to
expanded analysis with larger cohorts. Nonetheless, the class-specific
metastatic pathway enrichment we observe (in **table 2**) could be a result of
the pathways directly promoting metastatic phenotypes in these classes, or
because they acquire a stronger or more complete oncogenic phenotype, since we
don’t know the order in which they were acquired in individual tumor samples.
The fact that most metastatic enrichment is cluster-specific and has a tendency
to overlap with cluster-specific pathways from non-metastatic tumors of
neighboring clusters favors the latter interpretation, and suggests a mechanism
of complementarity, in which newly acquired mutations result in similar classes
converging on a single deadly phenotype that captures most or all of the
critical features of end-stage cancer. Some clusters were characterized by
cluster-specific pathways that seem superficially unrelated; for example cluster
3 has mutations impacting protein translation and turnover, whereas cluster 5
has defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
traditional extra-cellular signaling pathways (as exemplified in **table 1**).
It may be that the primary effects of mutations in protein and RNA metabolism in
clusters 3 and 5 point to currently unappreciated mechanisms impacting hallmark
pathways. Clusters 3 and 5 are also marked by a decidedly different set of
pathways in metastasis. Clusters 3 and 5 notably have (in addition to signaling
pathways) mutations in glucagon and glycogen metabolism, respectively, in their
metastases (**table 2**), consistent with energy metabolism as one of the new
hallmarks [@hallmarks_II], but also distinguishing these classes from the others
in the order or nature of hallmark acquisition. Thus, even at relatively low
power, our analysis of metastasis further supports pathway-centric analysis of
tumors as indicative of real differences in underlying biology. As a major
caveat to this, we acknowledge that many therapies do target specific signaling
pathways ( _e.g._ EGFR or HER2), but understanding the broader context of the
genetic background and pathway vulnerability of tumors containing such markers
may aid in creating smarter combination therapies. We submit that when we
discover the requirements of each cluster with respect to pathway disruptions
and metastasis we may be able to target them therapeutically and prevent further
adaptation.
